---
title: "COVID-19 patients' blood samples analysis"
author: "Aleksandra Mizera"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  html_document: 
    toc: yes
    toc_float: yes
    theme: cosmo
    df_print: kable
---

# Executive summary
<!-- TODO  -->

# Environment preparation
Setting global options and initialization of the seed variable, which is necessary for ensuring reproducibility.
```{r setup}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(message = FALSE)

set.seed(23)
```

Loading the libraries used in the project.
```{r libs}
library(knitr)
library(xlsx)
library(dplyr)
library(tidyr)
library(ggplot2)
```

Loading dataset into dataframe without any pre clean-up with cache option checked.
```{r load_data, cache=TRUE}
path <- './res/wuhan_blood_sample_data_Jan_Feb_2020.xlsx'
df_raw <- read.xlsx(path, sheetIndex=1, header=TRUE)
```


# Dataset description
<!-- TODO: general information about dataset origin, references -->

## General information
<!-- dataframe dimentions, cols: name, type, units -->

## Basic statistics
<!-- TODO: df summary -->

# Individual attributes' value analysis

# Correlation between attributes

# Change of attribute value over time

# Survival prediction

# Attribute importance analysis

# Conclusion

