---
title: "COVID-19 patients' blood samples analysis"
author: "Aleksandra Mizera"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  html_document: 
    toc: yes
    toc_float: yes
    df_print: kable
    theme: lumen
---

# Executive summary
<!-- TODO  -->

# Environment preparation
Setting global options and initialization of the seed variable, which is necessary for ensuring reproducibility.
```{r setup}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(message = FALSE)

set.seed(23)
```

Loading the libraries used in the project.
```{r libs}
library(knitr)
library(xlsx)
library(lubridate)
library(dplyr)
library(tidyr)
library(arules)
library(DT)
library(kableExtra)
library(plotly)
library(ggplot2)
library(corrplot)
library(codebook)
```

# Dataset processing
The dataset is based on the medical records of 375 patients suffering from COVID-19 collected between 10 January and 18 February 2020 in Tongji Hospital in China. The medical records were gathered by using standard case report forms, which included clinical, epidemiological, demographic, laboratory and mortality outcome information. Data collecting methods and some basic statistics are presented in [Tan et al][2] article published in Nature, 14 May 2020.
\
\
This section covers loading source data, initial remarks about its content, cleaning phase and detailed dataset description.

## Loading source data
The dataset can be downloaded under [this][1] link. In this project it's stored in a local subdirectory to ensure availability, even when it's deleted from the original source.
\
\
Loading source data into a dataframe without any pre clean-up with caching option checked.
```{r load_data, cache=TRUE}
path <- '../res/wuhan_blood_sample_data_Jan_Feb_2020.xlsx'
df_raw <- read.xlsx(path, sheetIndex=1, header=TRUE)
```

## Initial remarks and pre-cleanup
Raw dataset counts 6120 rows and 81 columns. Each row has a timestamp when the blood test result was released, which is stored in RE_DATE column. As we notice, values in this column are not unique and there are some missing values. Looking at the arrangement of the rows we can conclude that the dataset is ordered by the increasing value of patient ID (PATIENT_ID column). What's more, multiple lines are linked to a single patient, but only the first one is properly tagged with patient ID.
```{r initial_info, echo=FALSE}
data.frame(rows = nrow(df_raw), cols = ncol(df_raw), dates = length(unique(df_raw$RE_DATE)), dates_na = sum(is.na(df_raw$RE_DATE)), id_na = sum(is.na(df_raw$PATIENT_ID))) %>%
  kbl(col.names = c('rows', 'columns','unique RE_DATE', 'NAs in RE_DATE', 'NAs in PATIENT_ID')) %>%
  kable_styling() 
```

### Unique keys
In order to get a distinctive key for each line we will combine RE_DATE with PATIENT_ID values. In the first step we will remove rows with NA in RE_DATE, because after further checking we can see that all of the blood test results columns are empty for these lines. Then we will fill the PATIENT_ID column.
```{r get_unique_keys}
all(sapply(df_raw[is.na(df_raw$RE_DATE), c(8:81)], is.na))

df <- df_raw %>% 
  filter(!is.na(RE_DATE)) %>%
  fill(PATIENT_ID)

nrow(df)
length(unique(paste(df$PATIENT_ID, df$RE_DATE)))
```

### Attribute names and data types
The columns can be divided into 2 categories:

* information about a patient
* specific blood test result

Short notes about the first 7 columns, containing information about a patient, are shown below.
```{r personal, echo=FALSE}
str(df[,1:7])
```
As we can see, the gender and outcome of treatment columns are encoded with numbers. Both of these attributes have only 2 unique values, so we will transform them into factors for more clarity. The information about encoding was found in the article mentioned before, which was used as a reference in this analysis.
```{r as_factors}
df <- df %>%
  mutate(gender = factor(gender, labels = c('male', 'female')),
         outcome = factor(outcome, labels = c('survived', 'deceased')))
```

The next table presents a summary for the remaining columns, which store results of specific blood tests.
```{r blood_tests, echo=FALSE}
summary(df[,8:81]) %>%
  kbl() %>%
  kable_styling(bootstrap_options = c("striped", "hover")) %>%
  kableExtra::scroll_box(width = "100%")
```

The author of the original dataset decided to label columns with full-length names of the blood tests, which are too long and incomprehensible for our purposes. Also in some of names occur special symbols, which were degenerated during data loading. To obtain easy to work with column names the following steps were performed:

* replacing too long names with its abbreviations used in medical jargon (if one exists) or made-up short forms
* standardizing separator symbol to '.'
* replacing string '...', which is a remaining of deformed '(%)' from loading data, with '.percent'
* removing preceding and trailing redundant separator symbol occurrences
* converting multi-word names to lower case with an exception to names changed by hand
```{r rename}
df <- df %>%
  rename("hs.cTnI" = "Hypersensitive.cardiac.troponinI") %>%
  rename("PT" = "Prothrombin.time") %>%
  rename("IL.2.receptor" = "Interleukin.2.receptor") %>%
  rename("ALP" = "Alkaline.phosphatase") %>%
  rename("IL.10" = "Interleukin.10") %>%
  rename("AT" = "antithrombin") %>%
  rename("IL.8" = "Interleukin.8") %>%
  rename("RDW" = "Red.blood.cell.distribution.width.") %>%
  rename("q.anti.TP" = "Quantification.of.Treponema.pallidum.antibodies") %>%
  rename("MCV" = "mean.corpuscular.volume") %>%
  rename("WBC.count" = "White.blood.cell.count") %>%
  rename("TNF.alpha" = "Tumor.necrosis.factorα") %>%
  rename("MCHC" = "mean.corpuscular.hemoglobin.concentration") %>%
  rename("IL.1.beta" = "Interleukin.1β") %>%
  rename("RBC.count" = "Red.blood.cell.count") %>%
  rename("MPV" = "Mean.platelet.volume") %>%
  rename("RBC.DW.SD" = "RBC.distribution.width.SD") %>%
  rename("TT" = "Thrombin.time") %>%
  rename("q.anti.HCV" = "HCV.antibody.quantification") %>%
  rename("D.Dimer" = "D.D.dimer") %>%
  rename("AST" = "aspartate.aminotransferase") %>%
  rename("NT.proBNP" = "Amino.terminal.brain.natriuretic.peptide.precursor.NT.proBNP.") %>%
  rename("LDH" = "Lactate.dehydrogenase") %>%
  rename("P.LCR" = "platelet.large.cell.ratio.") %>%
  rename("IL.6" = "Interleukin.6") %>%
  rename("FDP" = "Fibrin.degradation.products") %>%
  rename("PDW" = "PLT.distribution.width") %>%
  rename("GGT" = "γ.glutamyl.transpeptidase") %>%
  rename("INR" = "International.standard.ratio") %>%
  rename("X2019.nCoV.NAT" = "X2019.nCoV.nucleic.acid.detection") %>%
  rename("MCH" = "mean.corpuscular.hemoglobin.") %>%
  rename("aPTT" = "Activation.of.partial.thromboplastin.time") %>%
  rename("hs.CRP" = "High.sensitivity.C.reactive.protein") %>%
  rename("q.anti.HIV" = "HIV.antibody.quantification") %>%
  rename("ALAT" = "glutamic.pyruvic.transaminase") %>%
  rename_with(~gsub('_', '.', .)) %>%
  rename_with(~gsub('\\.{3}$', '.percent', .)) %>%
  rename_with(~gsub('^X\\.{3}|\\.$', '', .)) %>%
  rename_with(tolower, matches("^[[:upper:]]{1}[[:lower:]]+|^[[:upper:]]{2,}\\.{1}[[:upper:]]{2,}$", ignore.case = FALSE))
```

Obtained names are presented below.
```{r cleaned_names, echo=FALSE}
colnames(df)
```

All of these attributes are of numeric type. In the summary we could observe that in one column some values are equal to -1. According to the article, this number was selected as a label for missing values. After checking the whole content of that column, we can see that there are only two different values {NA, -1}, which indicates that it's filled only with missing values, thus should be removed from the dataset.
```{r show_empty_col, echo=FALSE}
colnames(df)[apply(df, 2, function(x) any(x == -1, na.rm = TRUE))]
unique(df[, colnames(df)[apply(df, 2, function(x) any(x == -1, na.rm = TRUE))]])
```

```{r delete_empty_col, echo=TRUE}
df <- df %>% select(-c(colnames(df)[apply(df, 2, function(x) any(x == -1, na.rm = TRUE))]))
```

## Groups detection
The number of lines per patient is clearly inconstant, which suggests that every patient could have undergone various combinations of specific blood tests. Also, it's common knowledge that one sample of drawn blood is enough to perform multiple tests. We would like to check if some of the blood tests available in our dataset were more likely to be performed together.
\
\
Firstly, let's check how many unique blood tests combinations are present in the dataset.
```{r check_trans}
trans <- unique(apply(!is.na(df[, 8:80]), 1, which))
length(trans)
```

Now we will try to find closed frequent itemsets with apriori algorithm. The maximum number of items per itemset is set to 25. The maximum subset checking time is 60 seconds, support = 0.09 and confidence = 0.8.

```{r apriori}
freq_itemsets <- apriori(trans, parameter = list(support=0.09, maxlen=25, maxtime=60, target="closed frequent itemsets"))

max_freq_itemsets <- freq_itemsets[is.maximal(freq_itemsets)]
itemLabels(max_freq_itemsets) <- colnames(df[, 8:80])
inspect(sort(max_freq_itemsets, by='support'))
```
The algorithm returned 12 closed frequent itemsets. Analyzing the statistics above, we can draw some conclusions:

1. Itemset 3 contains tests detecting presence of antibodies HIV, HCV, Treponema Pallidum antibodies and HBsAg protein (arises in presence of HBV). The aim of these tests is to determine if the patient is a carrier of any of those pathogens.
2. Itemsets 4, 11 and 12 differentiate from each other at most with 3 elements, thus their intersection can be treated as a single itemset.
3. Itemsets 4, 11 and 12 consist mostly of items found in standard Blood Biochemistry tests.
4. Itemset 7 consists of items found in the standard Complete Blood Count test.
5. The items in itemset 5 are known as coagulation tests, which measure blood’s ability to clot.
6. Glucose and procalcitonin levels are often checked together as indicated by itemset 8. Procalcitonin level rises in response to a pro-inflammatory stimulus, especially of bacterial origin. It is therefore often classed as an acute phase reactant.

## Final cleanup
For now the percent of NA values in the dataset is too high.
```{r na_percent}
(sum(is.na(df)) / (nrow(df) * ncol(df))) * 100
```

To minimize the number of missing values, we will collapse all rows linked to one patient with tests' results, which were released on the same day. In case of multiple results of the same test during one day, the latest will be saved.
```{r collapse_rows}
df_clean <- df %>%
  mutate(re.date = floor_date(re.date, unit = "day")) %>%
  group_by(patient.id, re.date) %>%
  summarize(across(age:outcome, last), across(hs.cTnI:creatinine, ~.[last(which(!is.na(.)))])) %>%
  ungroup()
```

Let's remove patients who have not been tested for at least 50% of all available tests.
```{r remove_patients}
ids <- df_clean %>%
  group_by(patient.id) %>%
  summarize(across(hs.cTnI:creatinine, ~any(!is.na(.)))) %>%
  summarize(patient.id = patient.id, n = rowSums(select(., c(hs.cTnI:creatinine)))) %>%
  ungroup() %>%
  filter(n > ncol(select(df_clean, c(hs.cTnI:creatinine))) * 0.5)

df_clean <- df_clean %>%
  filter(patient.id %in% ids$patient.id)

(sum(is.na(df_clean)) / (nrow(df_clean) * ncol(df_clean))) * 100
```
As we can see, the percentage of NA values decreased from approximately 80% to 51%, which is still high, but deleting even more records would significantly reduce the size of the dataset.

# Dataset description
Final version of the dataset contains 1698 rows and 80 columns. Each row represents a set of blood tests results, which were performed for a single patient on a given day.
```{r dim_clean}
dim(df_clean)
```

Detailed description of individual columns of the dataset with some basic statistics are introduced in the codebook below.
```{r codebook, echo=FALSE}
label_unit <- list(
              patient.id = c("patient's identifier number", NA),
              re.date = c("date of blood result release", NA),
              age = c("patient's age", NA),
              gender = c("patient's gender", NA),
              admission.time = c("date of patient's admission to the hospital", NA),
              discharge.time = c("date of patient's discharge from the hospital", NA),
              outcome = c("outcome of treatment", NA),
              hs.cTnI = c("level of cardiac troponin I proteins", "ng/L"),
              hemoglobin = c("amount of hemoglobin", "g/L"),
              serum.chloride = c("amount of chloride", "mmol/L"),
              PT = c("time required for blood to clot", "seconds"),
              procalcitonin = c("level of procalcitonin", "ng/mL"),
              eosinophils.percent = c("percentage of eosinophils in white blood cells", "%"),
              IL.2.receptor = c("level of interleukin-2 receptor", "pg/mL"),
              ALP = c("level of alkaline phosphatase", "U/L"),
              albumin = c("amount of albumin protein", "g/L"),
              basophil.percent = c("percentage of basophils in white blood cells", "%"),
              IL.10 = c("level of serum IL-10", "pg/mL"),
              total.bilirubin = c("total amount of a substance called bilirubin (directed + indirected)", "mmol/L"),
              platelet.count = c("number of platelets in a volume of blood", "10^9/L"),
              monocytes.percent = c("percentage of monocytes in white blood cells", "%"),
              AT = c("activity and the amount of antithrombin", "%"),
              IL.8 = c("level of serum IL-8", "pg/mL"),
              indirect.bilirubin = c("total bilirubin minus the direct bilirubin levels in the bloodstream", "mmol/L"),
              RDW = c("amount of red blood cell variation in volume and size", "%"),
              neutrophils.percent = c("percentage of neutrophils in white blood cells", "%"),
              total.protein = c("total amount of albumin and globulin in blood", "g/L"),
              q.anti.TP = c("presence of Treponema Pallidum antibodies in blood", NA),
              prothrombin.activity = c(NA, NA),
              HBsAg = c("Hepatitis B surface antigen", "IU/mL"),
              MCV = c("average size of erythrocytes", "fL"),
              hematocrit = c("volume percentage of red blood cells (RBC) in blood", "%"),
              WBC.count = c("number of white blood cells in a volume of blood", "10^9/L"),
              TNF.alpha = c("marker associated with an inflammatory response", "pg/mL"),
              MCHC = c("average concentration of hemoglobin inside a single red blood cell", "g/L"),
              fibrinogen = c("level of fibrinogen", "g/L"),
              IL.1.beta = c("level of interleukin-1-β", "pg/mL"),
              urea = c("amount of urea nitrogen found in blood", "mg/dL"),
              lymphocyte.count = c("number of lymphocytes in a volume of blood", "10^9/L"),
              PH.value = c("degree of acidity or alkalinity", "[0:14]"),
              RBC.count = c("number of red blood cells in a volume of blood", "	10^12/L"),
              eosinophil.count = c("number of eosinophils in a volume of blood", "10^9/L"),
              corrected.calcium = c("an estimate of the total calcium concentration", "mmol/L"),
              serum.potassium = c("amount of potassium", "mmol/L"),
              glucose = c("blood glucose level", "mmol/L"),
              neutrophils.count = c("number of neutrophils in a volume of blood", "10^9/L"),
              direct.bilirubin = c("amount of conjugated bilirubin in blood", "mmol/L"),
              MPV = c("average size of platelets", "fL"),
              ferritin = c("blood ferritin level", "ng/L"),
              RBC.DW.SD = c("measure of the range of variation of red blood cell (RBC) volume", "fL"),
              TT = c("time required for a fibrin clot to form following the addition of thrombin to plasma", "seconds"),
              lymphocyte = c("percentage of lymphocytes in white blood cells", "%"),
              q.anti.HCV = c("presence of HCV antibodies in blood", NA),
              D.Dimer = c("level of D-Dimer in blood (blood clots detection)", "mg/L"),
              total.cholesterol = c("amount of cholesterol and triglycerides in blood", "mmol/L"),
              AST = c("level of aspartate aminotransferase (ALT, AST) enzyme in blood", "U/L"),
              uric.acid = c("amount of uric acid in blood", "µmol/L"),
              HCO3 = c("amount of bicarbonate (HCO3-) in blood", "mmol/L"),
              calcium = c("measurement of total calcium concentration", "mmol/L"),
              NT.proBNP = c("level of NT-proB-type Natriuretic Peptide (NT-proBNP) in blood", "ng/L"),
              LDH = c("level of lactate dehydrogenase (LDH) in blood", "U/L"),
              P.LCR = c("the proportion of platelets greater than 12 fL (P-LCR)", "%"),
              IL.6 = c("level of interleukin-6 protein in blood", "pg/mL"),
              FDP = c("amount of fibrin degradation products (FDPs) in blood", "mg/L"),
              monocytes.count = c("number of monocytes in a volume of blood", "10^9/L"),
              PDW = c("variation of platelet size distribution (PDW)", "%"),
              globulin = c("amount of globulin protein", "g/L"),
              GGT = c("level of Gamma-Glutamyl Transferase (GGT)", "U/L"),
              INR = c("calculation based on results of a Prothrombin Time test concerning time required for blood clotting", "ratio"),
              basophil.count = c("number of basophils in a volume of blood", "10^9/L"),
              MCH = c("average quantity of hemoglobin present in a single red blood cell (MCH)", "pg"),
              aPTT = c("time required for blood to clot (aPTT)", "seconds"),
              hs.CRP = c("more sensitive version of CRP test determining level of C-Reactive Protein in blood", "mg/L"),
              q.anti.HIV = c("presence of HIV antibodies in blood", NA),
              serum.sodium = c("amount of sodium", "mmol/L"),
              thrombocytocrit = c("thrombocyte (platelet) content of blood", NA),
              ESR = c("Erythrocyte Sedimentation Rate test", "mm/hr"),
              ALAT = c("level of Alanine transaminase enzyme in blood", "U/L"),
              eGFR = c("estimated Glomerular Filtration Rate", "mL/min/1.73m^2"),
              creatinine = c("level of creatinine", "μmol/L"))

codebook <- as(df_clean, "data.frame")

for (x in seq_along(label_unit)) {
  attr(codebook[, x], "label") <- label_unit[[x]][1]
  attr(codebook[, x], "units") <- label_unit[[x]][2]
}

datatable(codebook_table(codebook),
    options = list(scrollX = TRUE,
                   autoWidth = TRUE,
                   paging = TRUE,
                   pageLength = 10,
                   lengthMenu = c(5, 10, 15, 20),
                   columnDefs = list(list(className = 'dt-right', targets = '_all'),
                                     list(width = '300px', targets = 2),
                                     list(width = '150px', targets = c(10:12)))))
```

## Sample

A sample of cleaned data is presented below.
```{r sample, echo=FALSE}
sample <- head(df_clean, 10)
sample <- sample %>%
  mutate_if(is.POSIXct, as.character)

datatable(sample,
          options = list(scrollX = TRUE,
                         autoWidth = TRUE,
                         paging = TRUE,
                         columnDefs = list(list(className = 'dt-right', targets = '_all'),
                                           list(width = '150px', targets = c(2, 5, 6))),
                         dom = 't'))
```

# Dataset analysis
This section is dedicated to attribute values analysis and visualization. It covers not only separate study of each variable but also attempts of finding correlation between them.

## Data over time

The following chart presents how many blood tests of a given type were performed during each day from January 10 to February 18, 2020.
```{r fig_tests_per_day, fig.width=8, fig.height=6, echo=FALSE, warning=FALSE}
margins <- list(t = 100, pad = 4)

df_clean %>%
  group_by(re.date) %>%
  summarize(across(hs.cTnI:creatinine, ~sum(!is.na(.)))) %>%
  gather(variable, n, hs.cTnI:creatinine, factor_key = TRUE) %>%
  plot_ly(x=~re.date, y=~n, type="scatter", color=~variable, mode="markers") %>%
  layout(title = 'Tests by type per day',
         hovermode = 'compare',
         margin = margins,
         xaxis = list(tickangle = 90,
                      dtick = 3*86400000),
         yaxis = list(tick0 = 0, dtick = 10))
```
At first glance, we observe a significant rise in the total number of tests after January 27, which suggests that from this day forward more people infected with COVID-19 were admitted to the hospital on a daily basis. Another fact is that before that day the variety of blood test types, which were performed, was visibly lower. Throughout the whole considered period, the most frequently carried out exams were tests included in Complete Blood Count, Blood Biochemistry and glucose level categories.

To confirm the suspicion of a growing number of admitted patients starting on January 27, let's construct a chart visualizing the number of admitted and discharged patients on a given day.
```{r fig_admissions_discharges, fig.width=8, fig.height=6, echo=FALSE, warning=FALSE}
df_clean %>%
  mutate_if(is.POSIXct, ~floor_date(., unit = "day")) %>%
  group_by(patient.id) %>%
  summarize(across(admission.time:discharge.time, last)) %>%
  gather(type, date, admission.time:discharge.time, factor_key = TRUE) %>%
  group_by(type,date) %>%
  summarise(n = n()) %>%
  plot_ly(x=~date, y=~n, type="scatter", color=~type, mode="lines+markers") %>%
  layout(title = 'Admissions and discharges per day',
         hovermode = 'compare',
         margin = margins,
         xaxis = list(tickangle = 90,
                      dtick = 3*86400000))
```
The number of patients admitted to the hospital on a single day peaked on February 10, whereas the highest number of discharged patients was observed on February 18. 

## Attribute analysis
This section is broke into 3 parts: analyzing data concerning patient's personal information, blood test results for each type and correlation between them.

### Information about a patient

The following overlaying histograms portray age distribution of COVID-19 survivors and victims.
```{r fig_age, fig.width=6, fig.height=4, echo=FALSE, warning=FALSE}
survived<- df_clean %>%
  group_by(patient.id) %>%
  summarize(across(c(age, outcome), last)) %>%
  filter(outcome == 'survived') %>%
  count(age)

deceased <- df_clean %>%
  group_by(patient.id) %>%
  summarize(across(c(age, outcome), last)) %>%
  filter(outcome == 'deceased') %>%
  count(age)

plot_ly(alpha = 0.6) %>%
  add_histogram(x = ~rep(survived$age, survived$n), name = 'survived') %>%
  add_histogram(x = ~rep(deceased$age, deceased$n), name = 'deceased') %>%
  layout(barmode = 'overlay',
         title = 'Age distribution',
         hovermode = 'compare',
         margin = margins,
         xaxis = list(title = 'age bins', range = c(10, 100), tickvals = seq(20, 100, 10)),
         yaxis = list(title = 'n'))
```
The majority of patients, who died for discussed disease, were people aged 65 years or older. Only two young people in age ranges <20 and 30-34 were an exception from this trend. Reported survivors were mainly people under 70 years.
\
\
The next figure shows gender and outcome of treatment distributions.
```{r fig_gender_outcome, fig.width=6, fig.height=4, echo=FALSE, warning=FALSE}
fig_sex <- df_clean %>%
  group_by(patient.id) %>%
  summarize(gender = last(gender)) %>%
  count(gender) %>%
  plot_ly(x = ~gender, y = ~n, color = ~gender, type = 'bar') %>%
  layout(margin = margins,
         xaxis = list(title = "gender"))

fig_outcome <- df_clean %>%
  group_by(patient.id) %>%
  summarize(outcome = last(outcome)) %>%
  count(outcome) %>%
  plot_ly(x = ~outcome, y = ~n, color = ~outcome, colors = "Dark2", type = 'bar') %>%
  layout(margin = margins,
         xaxis = list(title = "outcome of treatment"))

subplot(fig_sex, fig_outcome, shareY = T, titleX = T) %>%
  layout(title = "Age and outcome distribution",
         showlegend = F)
```
In a set of 357 patients, 210 of them were male and 147 were female. The ratio of the deceased patients to the ones, who were cured, equals to approximately 0.84, which even at the time was alarmingly high.
\
\
Let's calculate the total amount of time spent in the hospital by each patient and group these values by the outcome of treatment.
```{r fig_followup_time, fig.width=6, fig.height=4, echo=FALSE, warning=FALSE}
df_clean %>%
  group_by(patient.id) %>%
  summarize(across(c(admission.time:outcome), last)) %>%
  mutate(followup.time = difftime(discharge.time, admission.time, units = "days")) %>%
  select(c(followup.time, outcome)) %>%
  plot_ly(y = ~followup.time[which(outcome == 'survived')], type = 'box', name = 'survived') %>%
  add_trace(y = ~followup.time[which(outcome == 'deceased')], type = 'box', name = 'deceased') %>%
  layout(title = 'Follow-up time by outcome',
         margin = margins,
         yaxis = list(title = 'time [days]'))
```
The figure shows that the mean number of days of medical follow-up for the deceased patients was less than the corresponding value for the survivors. The conclusion is that people who were reported dead were admitted to the hospital mostly in already severe condition, whereas patients who got the treatment in early onset of disease had better chances of getting cured.

### Blood tests
Based on the assumption that potential readers of this analysis are not proficient in interpreting blood test results, calculating basic statistics for each column won't be very helpful in understanding what they mean. To avoid this, the data in each column was grouped by the outcome of treatment. Comparing results of these two classes gives us a useful point of view in deciding, which value ranges can be considered abnormal for a specific variable.
\
```{r fig_tests, fig.width=10, fig.height=50, echo=FALSE}
df_clean %>%
  select(outcome:creatinine) %>%
  gather(variable, value, hs.cTnI:creatinine, factor_key = TRUE, na.rm = T) %>%
  ggplot() +
    facet_wrap(~variable, ncol = 4, scales = "free") +
    geom_boxplot(mapping=aes(x=outcome, y=value, color=outcome))
```
\
TODO

## Correlation
In order to check if some of variables with blood test results values are correlated, we calculate Pearson correlation coefficient for columns [8:80] obtaining correlation matrix. Beside the value of the correlation coefficient we will take under consideration significance level as well. To do so, we calculate p-values and store them in another matrix.

The plot below presents a correlogram combined with the significance test sorted by the coefficient value. The diagonal is omitted for better readability.
```{r corr, fig.width=10, fig.height=10, echo=FALSE}
M = cor(df_clean[, 8:80], method = "pearson", use = "complete.obs")

cor.mtest <- function(mat, ...) {
    mat <- as.matrix(mat)
    n <- ncol(mat)
    p.mat<- matrix(NA, n, n)
    diag(p.mat) <- 0
    for (i in 1:(n - 1)) {
        for (j in (i + 1):n) {
            tmp <- cor.test(mat[, i], mat[, j], ...)
            p.mat[i, j] <- p.mat[j, i] <- tmp$p.value
        }
    }
  colnames(p.mat) <- rownames(p.mat) <- colnames(mat)
  p.mat
}

p_mat <- cor.mtest(df_clean[, 8:80], method = "pearson", use = "complete.obs")

corrplot(M, type="upper", order="hclust", p.mat = p_mat, sig.level = 0.05, insig = "blank", tl.col="black", tl.cex = 0.7, diag = F)
```
In the above figure, correlations with p-value > 5% are considered as insignificant. In this case the correlation coefficient values are left blank.
\
\
We observe on the correlogram the following facts:
\
TODO

# Classifier
This section describes preprocessing of the dataset and build, training and validation of the model for a classifier.

## Dataset preprocessing
TODO

## Training
TODO

## Attribute importance
TODO

# Conclusion
TODO



<!-- links -->
[1]: http://www.cs.put.poznan.pl/dbrzezinski/teaching/zed/wuhan_blood_sample_data_Jan_Feb_2020.xlsx "source data"
[2]: https://www.nature.com/articles/s42256-020-0180-7 "article"