---
title: "COVID-19 patients' blood samples analysis"
author: "Aleksandra Mizera"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  html_document: 
    toc: yes
    toc_float: yes
    df_print: kable
    theme: lumen
    highlight: tango
---

# Executive summary
<!-- TODO  -->

# Environment preparation
Setting global options and initialization of the seed variable, which is necessary for ensuring reproducibility.
```{r setup}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(message = FALSE)

set.seed(23)
```

Loading the libraries used in the project.
```{r libs}
library(knitr)
library(xlsx)
library(lubridate)
library(dplyr)
library(tidyr)
library(ggplot2)
```

Loading dataset into dataframe without any pre clean-up with caching option checked.
```{r load_data, cache=TRUE}
path <- '../res/wuhan_blood_sample_data_Jan_Feb_2020.xlsx'
df_raw <- read.xlsx(path, sheetIndex=1, header=TRUE)
```

Cleaning data consisted of the following steps:

1. Normalization of column names
    + standardizing separator symbol to '.'
    + converting to lower case
    + removing preceding and trailing redundant separator symbol occurrences
    + shortening too long variable names
2. Conversion of selected columns' data types
    + setting column 'gender' as factor with 2 levels {1: male, 2: female}
    + setting column 'outcome' as factor with 2 levels {0: survived, 1: deceased}
3. Elimination of redundant columns
    + removing column 're.date' containing the date of the blood test
    + transforming 2 columns with admission and discharge dates of the patient into one column containing the amount of time spent in hospital
4. Collapse of multiple rows connected to one patient into single observation
    + grouping data by follow.up.time column after ensuring the uniqueness of its values
    + calculating the mean value for numerical attributes concerning information about blood samples (columns 5-79)
5. Decision on missing values
    + NaN values were replaced with ...
  
```{r clear_data}
df_clean <- df_raw %>%
  rename_with(~tolower(gsub('_', '.', .))) %>%
  rename_with(~gsub('^x\\.{3}|\\.{1,3}$', '', .)) %>%
  rename(nt.probnp = amino.terminal.brain.natriuretic.peptide.precursor.nt.probnp) %>%
  mutate(gender = factor(gender, labels = c('male', 'female')),
         outcome = factor(outcome, labels = c('survived', 'deceased')),
         admission.time = difftime(discharge.time, admission.time, units = "auto")) %>%
  select(-c(re.date, discharge.time)) %>%
  rename(follow.up.time = admission.time) %>%
  group_by(follow.up.time) %>%
  summarise(across(0:4, first), across(5:(ncol(.)-1), mean, na.rm = TRUE)) %>%
  ungroup()
```

A sample of cleaned data is presented below.

# Dataset description
<!-- TODO: general information about dataset origin, references -->


## General information
<!-- dataframe dimentions, cols: name, type, units -->

## Basic statistics
<!-- TODO: df summary -->

# Individual attributes' value analysis

# Correlation between attributes

# Change of attribute value over time

# Survival prediction

# Attribute importance analysis

# Conclusion

